### 2018-05-05 papers/synth/encoding.tex
some cleanup
asd

%
In this section, we will present a list of synthesis problems that may
arise from partial available information about a VTS and our solution
for the problems.

@@

In our synthesis method, we take a VTS $G =
(\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$ as input.
%
We allow activity functions not to be specified.
%
We construct the following constraints to encode the available information
about $G$.
%
\begin{align*}
  \texttt{PresentE} &= \land \{e_{i,j,q,m}|(i,M_1,j),...,(i,M_{q'},j) \in \edges \land q \leq q' \land m \in M_q \}\\
  \texttt{PresentN} &= \land \{n_{i,m}| m \in \nlabel(i) \land i \in \nodes \}\\
  \texttt{PresentP} &= \land \{p_{i,j}| (i,j) \in \pairs \} \\
  \texttt{KnownActiveN} &= \land \{ a_{i,m} = \texttt{NodeFun}_m (n_{i,1},\dots,n_{i,\mu}) | \nodef_m \text{ is defined.} \} \\
  \texttt{KnownActiveE} &= \land \{ b_{i,j,q,m} = \texttt{EdgeFun}_m(e_{i,j,q,1}, .., e_{i,j,q,\mu})
                   | \edgef_m \text{ is defined.} \} \\
  \texttt{PresentCons} & = \texttt{PresentE} \land \texttt{PresentN} \land 
 \texttt{PresentP} \land \texttt{KnownActiveE} \land \texttt{KnownActiveN}
\end{align*}
We also collect the variables that are not set to true in $\texttt{PresentCons}$.
\begin{align*}
  \texttt{AbsentELabel} &=
  \{e_{i,j,q,m}|(i,M_1,j),...,(i,M_{q'},j) \in \edges \land 
                          0 < q \leq q' \land m \not\in M_{q} \}\\
  \texttt{AbsentE} &= \{e_{i,j,q}|(i,M_1,j),...,(i,M_{q'},j) \in \edges \land 
                    q' < q \leq \pi \}\\
  \texttt{AbsentNLabel} &= \{n_{i,m}| m \not\in \nlabel(i) \land i \in \nodes \}\\
  \texttt{AbsentP} &= \{p_{i,j}| (i,j) \not\in \pairs \} \\
  \texttt{UnnownActive} &=  \Land \{ a_{i,m} = \nodef_m (n_{i,1},\dots,n_{i,\mu}) | \nodef_m \text{ is undefined.} \} \union \\
   &\quad\quad\quad  \{ b_{i,j,q,m} = \edgef_m(e_{i,j,q,1}, .., e_{i,j,q,\mu})
                   | \edgef_m \text{ is undefined.} \}
\end{align*}
We have defined $\texttt{AbsentELabel}$, $\texttt{AbsentE}$, $\texttt{AbsentN}$, and
$\texttt{AbsentP}$
as sets.
%
They will be converted into formulas depending
on the different usage in the synthesis problems.

@@

We will do synthesis against the following property that says the VTS
is stable and 3-connected.
%
% The property is designed to balance the search space such that the synthesis procedure does not
% succeed with simply adding too many edges. 
%
\begin{align*}
  \texttt{Property} =  \texttt{Stability} \land \texttt{Connected}(3) 
  %\land \texttt{DisConnected}(4)
\end{align*}
The property was proposed in~\cite{plos}.
%
However, the biological relevance of the property is debatable and open for change.
%
Our tool is easily modifiable to support any other property that may be deemed 
interesting by the biologists.

@@

\begin{align*}
  &\texttt{Gate}( x, w_1, w_2 ) = ( x  = 1 \limplies w_1  \land w_2 ) \land  
  ( x = 2  \limplies w_1  \lor w_2 ) \land
  ( x = 3  \limplies w_1 ) \land ...\\
  &\texttt{Leaf}( x, [y_1,..,y_\mu] ) =
  \Land\limits_{l=1}^{\mu} ( x = 2l-1  \limplies y_{l}) \land ( x = 2l  \limplies \lnot y_{l})\\
  &\texttt{NNFTemplate}([z_1,..,z_{2\lambda}],[y_1,..,y_\mu] ) = \\
  &\exists w_1,..,w_{2\lambda}.\;w_1 \land \Land_{l=1}^{\lambda} w_{\lambda+l} = \texttt{Leaf}( z_{\lambda+l}, [y_1,..,y_\mu] ) \land
  w_{l} = \texttt{Gate}( z_l, w_{2l},w_{2l+1})
\end{align*}

Using the template we define the constraints $\texttt{FindFunctions}(z,\lambda)$
that encodes the candidate functions that satisfy the activity requirements,
where $z$ is the vectors of parameters for encoding parameters
for each molecule,
and $\lambda$ limits the size of the candidate functions. 
%
We fix the all other aspect of the VTS to be fixed via constraints
$\texttt{FixedForFunctions}$.
%
\begin{align*}
  &\texttt{FindFunctions}(z,\lambda) =\\
  &\Land \{ \bigwedge\limits_{i} n_{i,m} \limplies a_{i,m} = 
  \texttt{NNFTemplate}([z_{m,1},..,z_{m,2\lambda}],[n_{i,1},\dots,n_{i,\mu}] ) |
  \nodef_m \text{ is undefined}\} \\
  &\Land \{ \bigwedge\limits_{i,j,q} e_{i,j,q,m} \limplies b_{i,j,q,m} = 
  \texttt{NNFTemplate}([z_{i,j,q,1},..,z_{i,j,q,2\lambda}],[e_{i,j,q,1},..,e_{i,j,q,\mu}] ) | 
  \edgef_m \text{ is undefined}
  \}\\
  & \texttt{FixedForFunctions} = \texttt{PresentCons} \land\\
  & \quad \quad
\lnot \Lor \texttt{AbsentE} \;\union\; \texttt{AbsentELabel} \;\union\;
                    \texttt{AbsentNLabel} \;\union\;
                    \texttt{AbsentP}\\
  &\texttt{SynthFunction}(z,\lambda) = \\
  & \quad\quad
    (  \texttt{Consistancy}\land \texttt{Property} \land
   \texttt{FixedForFunctions} \land \texttt{FindFunctions}(slimit) )
\end{align*}
We construct $\texttt{SynthFunction}(z,\lambda)$ similar to the earlier
variations.
%
By reading of the values of $z$ in a satisfying model of the formula,
we learn the synthesized function.

\subsection{Fixing VTS by both adding/deleting parts:}
%
Now we will consider repairing of VTS by allowing not only addition but also
deletion of the molecules, edges, functions, or pairing matrix.
%
We have encoded the repairing in our tool by introducing flip bits
for each variable that is modifiable in the VTS.
%
We illustrate the repairing on one class of variables and rest can be
easily extended.
%
Let us consider repairing of node labels.
%
For each bit $n_{i,m}$, we create a bit $flip_{i,m}$.
%
We add constraints that take xor of VTS assigned values for  $n_{i,m}$
and $flip_{i,m}$.
%
We also limit how many times the $flip_{i,m}$ can be true, therefore
limiting the number of flips.
%
The above constraints are encoded in $\texttt{FlipN}(slimit)$.
\begin{align*}
    &\texttt{FlipN}(slimit,flip) = \Land \{n_{i,m}\lxor flip_{i,m}| m \in \nlabel(i) \land i \in \nodes \} \land \\
  & \quad \quad \quad
    \Land \{ \lnot n_{i,m}\lxor flip_{i,m}| m \not \in \nlabel(i) \land i \in \nodes \} \land  \sum\limits_{i,m} flip_{i,m} \leq slimit
\end{align*}
%
Similar to the earlier variations, we construct
$\texttt{SynthRepairNode}(slimit)$ for the repair.
%
In that, $\texttt{FixedForNodeRepair}$ encodes all the parts of VTS that do not change.
\begin{align*}
% \\ 
    & \texttt{FixedForNodeRepair} =  \texttt{PresentE} \land  
      \texttt{PresentP} \land \texttt{KnownActiveE} \land \texttt{KnownActiveN}
      \land \\
  & \quad \quad \texttt{UnknownActive} \land \lnot 
    \Lor \texttt{AbsentE} \;\union\; \texttt{AbsentNLabel} \;\union\;
    \texttt{AbsentP}\\
  &\texttt{SynthRepairNode}(slimit,flip) = \\
  & \quad\quad
    (  \texttt{Consistancy}\land \texttt{Property} \land
    \texttt{FixedForNodeRepair} \land \texttt{FlipN}(slimit,flip) )
\end{align*}
A satisfying model of $\texttt{SynthRepairNode}(slimit,flip)$ will assign some
$flip$ bits to true.
We will learn from the assignments the needed modifications in the VTS.

>>>>

%
In this section, we will present a list of synthesis problems that may
arise from the partial available information about a VTS and our synthesis method
for the problems.

@@

In our synthesis method, we take a VTS $G =
(\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$ as input.
%
We allow activity functions not to be specified.
%
We construct the following constraints to encode the available information
about $G$.
%
\begin{align*}
  \texttt{PresentE} &= \land \{e_{i,j,q,m}|(i,M_1,j),...,(i,M_{q'},j) \in \edges \land q \leq q' \land m \in M_q \}\\
  \texttt{PresentN} &= \land \{n_{i,m}| m \in \nlabel(i) \land i \in \nodes \}\\
  \texttt{PresentP} &= \land \{p_{i,j}| (i,j) \in \pairs \} \\
  \texttt{KnownActiveN} &= \land \{ a_{i,m} = \texttt{NodeFun}_m (n_{i,1},\dots,n_{i,\mu}) | \nodef_m \text{ is defined.} \} \\
  \texttt{KnownActiveE} &= \land \{ b_{i,j,q,m} = \texttt{EdgeFun}_m(e_{i,j,q,1}, .., e_{i,j,q,\mu})
                   | \edgef_m \text{ is defined.} \} \\
  \texttt{PresentCons} & = \texttt{PresentE} \land \texttt{PresentN} \land 
  \texttt{PresentP} \land \texttt{KnownActiveE} \land \\
  &  \texttt{KnownActiveN}
\end{align*}
We also collect the variables that are not set to true in $\texttt{PresentCons}$.
\begin{align*}
  \texttt{AbsentELabel} &=
  \{e_{i,j,q,m}|(i,M_1,j),...,(i,M_{q'},j) \in \edges \land 
                          0 < q \leq q' \land m \not\in M_{q} \}\\
  \texttt{AbsentE} &= \{e_{i,j,q}|(i,M_1,j),...,(i,M_{q'},j) \in \edges \land 
                    q' < q \leq \pi \}\\
  \texttt{AbsentNLabel} &= \{n_{i,m}| m \not\in \nlabel(i) \land i \in \nodes \}\\
  \texttt{AbsentP} &= \{p_{i,j}| (i,j) \not\in \pairs \} \\
  \texttt{UnnownActive} &=  \Land \{ a_{i,m} = \nodef_m (n_{i,1},\dots,n_{i,\mu}) | \nodef_m \text{ is undefined.} \} \union \\
   &\quad\quad\quad  \{ b_{i,j,q,m} = \edgef_m(e_{i,j,q,1}, .., e_{i,j,q,\mu})
                   | \edgef_m \text{ is undefined.} \}
\end{align*}
We have defined $\texttt{AbsentELabel}$, $\texttt{AbsentE}$, $\texttt{AbsentN}$, and
$\texttt{AbsentP}$
as sets.
%
They will be converted into formulas depending
on the different usage in the synthesis problems.

@@

We will do synthesis against the following property that says the VTS
is stable and 3-connected.
%
% The property is designed to balance the search space such that the synthesis procedure does not
% succeed with simply adding too many edges. 
%
\begin{align*}
  \texttt{Property} =  \texttt{Stability} \land \texttt{Connected}(3) 
  %\land \texttt{DisConnected}(4)
\end{align*}
The property was proposed in~\cite{shukla2017discovering}.
%
However, the biological relevance of the property is debatable and open for change.
%
Our tool is easily modifiable to support any other property that may be deemed 
interesting by the biologists.

@@

\begin{align*}
  &\texttt{Gate}( x, w_1, w_2 ) = ( x  = 1 \limplies w_1  \land w_2 ) \land  
  ( x = 2  \limplies w_1  \lor w_2 ) \\
  &\texttt{Leaf}( x, [y_1,..,y_\mu] ) =
  \Land\limits_{l=1}^{\mu} ( x = 2l-1  \limplies y_{l}) \land ( x = 2l  \limplies \lnot y_{l})\\
  &\texttt{NNFTemplate}([z_1,..,z_{2\lambda}],[y_1,..,y_\mu] ) = \\
  &\exists w_1,..,w_{2\lambda}.\;w_1 \land \Land_{l=1}^{\lambda} w_{\lambda+l} = \texttt{Leaf}( z_{\lambda+l}, [y_1,..,y_\mu] ) \land
  w_{l} = \texttt{Gate}( z_l, w_{2l},w_{2l+1})
\end{align*}

Using the template we define the constraints $\texttt{FindFunctions}(z,\lambda)$
that encodes the candidate functions that satisfy the activity requirements,
where $z$ is the vectors of parameters for encoding parameters
for each molecule,
and $\lambda$ limits the size of the candidate functions. 
%
We fix the all other aspect of the VTS to be fixed via constraints
$\texttt{FixedForFunctions}$.
%
\begin{align*}
  &\texttt{FindFunctions}(z,\lambda) =\\
  &\Land \{ \bigwedge\limits_{i} n_{i,m} \limplies a_{i,m} = 
  \texttt{NNFTemplate}([z_{m,1},..,z_{m,2\lambda}],[n_{i,1},\dots,n_{i,\mu}] ) \\
  & \hspace{8cm}|
  \nodef_m \text{ is undefined}\} \\
  &\Land \{ \bigwedge\limits_{i,j,q} e_{i,j,q,m} \limplies b_{i,j,q,m} = 
  \texttt{NNFTemplate}([z_{i,j,q,1},..,z_{i,j,q,2\lambda}],[e_{i,j,q,1},..,e_{i,j,q,\mu}] ) \\
  & \hspace{8cm} | \edgef_m \text{ is undefined}
  \}\\
  & \texttt{FixedForFunctions} = \texttt{PresentCons} \land\\
  & \quad \quad
\lnot \Lor \texttt{AbsentE} \;\union\; \texttt{AbsentELabel} \;\union\;
                    \texttt{AbsentNLabel} \;\union\;
                    \texttt{AbsentP}\\
  &\texttt{SynthFunction}(z,\lambda) = \\
  & \quad
    (  \texttt{Consistancy}\land \texttt{Property} \land
   \texttt{FixedForFunctions} \land \texttt{FindFunctions}(slimit) )
\end{align*}
We construct $\texttt{SynthFunction}(z,\lambda)$ similar to the earlier
variations.
%
By reading of the values of $z$ in a satisfying model of the formula,
we learn the synthesized function.

\subsection{Fixing VTS by both adding/deleting parts:}
%
Now we will consider repairing of VTS by allowing not only addition but also
deletion of the molecules, edges, functions, or pairing matrix.
%
We have encoded the repairing in our tool by introducing flip bits
for each variable that is modifiable in the VTS.
%
We illustrate the repairing on one class of variables and rest can be
easily extended.
%
Let us consider repairing of node labels.
%
For each bit $n_{i,m}$, we create a bit $flip_{i,m}$.
%
We add constraints that take xor of VTS assigned values for  $n_{i,m}$
and $flip_{i,m}$.
%
We also limit how many times the $flip_{i,m}$ can be true, therefore
limiting the number of flips.
%
The above constraints are encoded in $\texttt{FlipN}(slimit)$.
\begin{align*}
    &\texttt{FlipN}(slimit,flip) = \Land \{n_{i,m}\lxor flip_{i,m}| m \in \nlabel(i) \land i \in \nodes \} \land \\
  & \quad \quad \quad
    \Land \{ \lnot n_{i,m}\lxor flip_{i,m}| m \not \in \nlabel(i) \land i \in \nodes \} \land  \sum\limits_{i,m} flip_{i,m} \leq slimit
\end{align*}
%
Similar to the earlier variations, we construct
$\texttt{SynthRepairNode}(slimit)$ for the repair.
%
In that, $\texttt{FixedForNodeRepair}$ encodes all the parts of VTS that do not change.
\begin{align*}
% \\ 
    & \texttt{FixedForNodeRepair} =  \texttt{PresentE} \land  
      \texttt{PresentP} \land \texttt{KnownActiveE} 
      \land \\
  & \quad \texttt{UnknownActive} \land \texttt{KnownActiveN} \land \lnot 
    \Lor \texttt{AbsentE} \;\union\; \texttt{AbsentNLabel} \;\union\;
    \texttt{AbsentP}\\
  &\texttt{SynthRepairNode}(slimit,flip) = \\
  & \quad\quad
    (  \texttt{Consistancy}\land \texttt{Property} \land
    \texttt{FixedForNodeRepair} \land \texttt{FlipN}(slimit,flip) )
\end{align*}
A satisfying model of $\texttt{SynthRepairNode}(slimit,flip)$ will assign some
$flip$ bits to true.
We will learn from the assignments the needed modifications in the VTS.

### 2018-05-05 papers/synth/intro.tex
asd

We have implemented the encoding in a flexible tool,
which can handle a wide range of synthesis query.
%
We have applied our tool on two found-in-nature
VTSs~\cite{burri2004complete}.
%
In figure~\ref{fig:mukund-vts}, we present one of the two for
mammalian cells obtained by studying the
literature.\footnote{Please read appendix~\ref{sec:ex-vts} for more biological details.}
%
The VTS has six nodes and 20+ molecules.
%
We can easily check that stability condition is not satisfied for
many molecules for example ???.
%
It is clear that there are several unknowns about the VTS.
%
%
% Our tool 
% % and hollow arrows represent maturation edges. 
% We have used our tool to synthesize possible fixes to complete the
% graph.
% %
% Based on the variation we were able to synthesis the this graph
% within ??s.
%
Our synthesis tool reports that minimum ?? edges needs to be added to make
the VTS stable.
%
Our experiments suggest that the synthesis problems are
solvable by modern solvers and the synthesis technology may be useful
for biological research.
%
We have also applied our tool on various synthetic examples to
demonstrate that our tool scale upto the graph size size nodes, which
is the size of typical VTSs in a eukaryotic cell.

>>>>

We have implemented the encoding in a flexible tool,
which can handle a wide range of synthesis query.
%
We have applied our tool on several VTSs including two found-in-nature
VTSs.
%
In figure~\ref{fig:mukund-vts}, we present one of the two for
mammalian cells obtained by studying the
literature.\footnote{Please read appendix~\ref{sec:ex-vts} for more biological details.}
%
The VTS has six nodes and 55 molecules.
%
We can easily check that stability condition is not satisfied for
many molecules for example ???.
%
On the example, our synthesis tool reports that minimum ?? edges needed
to be added to make the VTS stable.
%
The dashed edges are the suggested edges from our tool.
%
Our experiments suggest that the synthesis problems are
solvable by modern solvers and the synthesis technology may be useful
for biological research.
%
% We have also applied our tool on various synthetic examples to
% demonstrate that our tool scale upto the graph size size nodes, which
% is the size of typical VTSs in a eukaryotic cell.
